---
title: "ING Lion's Den Pre-selection assignment"
author: "BifOfData - Jakub Sochacki, Michał Rejmak, Tomasz Kuczyński"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: pygments
---
## Introduction
In this report we are analyzing the historical data about the hard coal consumption data 
– monthly series per capita from July 1994 to December 2020 - 318 observations. Data has been collected by the National Statistic Office.

## Goal of the analysis  
**1)** Understand the data  
**2)** Time series decomposition  
**3)** Visualization of trends and seasonality  
**4)** Anomaly detection  
**5)** Examine stationarity and autocorrelation  
**6)** Selection of the best prediction method  
**6)** Forecast the hard coal consumption for 2021  

## Preparation
Libraries to be used:
```{r start, echo=TRUE, message=FALSE, warning = FALSE}
library(readxl)
library(rmdformats)
library(tidyverse)
library(forecast)
library(dplyr)
```
First step of analyzing time series data starts with a time plot.  
That is, the observations are plotted against the time of observation, with consecutive observations joined by straight lines.
```{r start1, echo=TRUE, message=FALSE, warning = FALSE}
Lions_Den_data = read_excel("Lions_Den_data.xlsx")
ggplot(Lions_Den_data, aes(x=DATE, y = `Hard coal consumption per capita [tones]`)) + geom_line() + ggtitle("Hard coal consumption 1994 - 2020")

```
  
Data preparation process for further analysis:  
data_ts is the time series object with frequency of 12 months and beginning in July 1994
```{r ts_preparation, echo=TRUE, message=FALSE}
data_ts <- ts(Lions_Den_data$`Hard coal consumption per capita [tones]`, frequency = 12, start = c(1994, 7))
```
As it is visible on the time series plot, there is a high and regular variability in hard coal consumption.  
To examine where it is coming from, it is imperative to analyze the seasonality of the time series.  
When investigating the seasonal plot of the data, it is easy to observe that the coal consumption changes over each year.  
Given that, it can be sated that the studied element depends on the season.  
It is specifically visible that the magnitude of the coal consumption increases July through February, and decreases February through July.  
```{r season_plot, echo=TRUE, message=FALSE, warning = FALSE}
ggseasonplot(data_ts, year.labels=TRUE, year.labels.left=TRUE, col=rainbow(n=25, start = 0, end = 24/25, alpha = 0.6)) +
  ylab("Hard coal consumption per capita [tones]") +
  ggtitle("Seasonal plot: Hard coal consumption")
```
```{r season_box}
boxplot(data_ts~cycle(data_ts))
```
Analogical data representation split into separate plots, aiming to better show the diminishing seasonality deviations:
```{r date_conversion, echo=TRUE, message=FALSE, warning = FALSE}
library(ggplot2)
data <- Lions_Den_data
data <- ?mutate(data, year = as.numeric(substring(DATE, 1, 4)), month = as.numeric(substring(DATE, 6, 7)), month_name = month.name[month])
# plots for each year
ggplot(data, aes(x=month, y = `Hard coal consumption per capita [tones]`)) + geom_line() + facet_wrap(~year)
```
Collection of plots portraying charcoal consumption volatility over the years, segregated by months.
Together with furhter analysis, discernible becomes the fact that the fluctuations in hard coal consumption
are notably less conspicuous during the summer period. Furthermore, there is also a matter of vital importance during
the year 2006 where major increase is observable.

```{r months}
# ggplot(data, aes(x=year, y = `Hard coal consumption per capita [tones]`)) + geom_line() + facet_wrap(~month)
```
## Time series decomposition
In order to properly analyze the time series, it is indispensable to perform a decomposition into three separate factors:  
**1.** Trend  
**2.** Seasonality  
**3.** Random / remainder factor  
Since the seasonal variation declines with the level of the series, a multiplicative model appears to be more accurate.  
Process of multiplicative time series decomposition:  
```{r decomposition}
# decomposition
data_decomp = decompose(data_ts, type="multiplicative")
plot(data_decomp)
```
Seasonally adjusted time series:
```{r season_removed}
 data_deseason <- data_ts / data_decomp$seasonal       # <<< tutaj sezonowość jest zawsze dodatnia
# plot(data_deseason)

autoplot(data_ts, series="Data") +
  autolayer(data_deseason, series="Seasonally adjusted") + ylab("Hard coal consumption per capita [tonnes]") + ggtitle("Seasonally adjusted hard coal consumption over years")
```

```{r czank2, echo=FALSE}

autoplot(data_ts, series="Data") +
  autolayer(trendcycle(stl(data_ts, s.window = "periodic")), series="Trend-cycle") + geom_smooth(method=lm, aes(colour="Linear trend fit"))+
  scale_colour_manual(values=c("red", "blue", "cyan3")) + ylab("Hard coal consumption per capita [tonnes]") + ggtitle("Trends of hard coal consumption over years")
```

```{r stl}
# stl decomposition
data_stl <- stl(data_ts, s.window = "periodic")
plot(data_stl)

# deseasoned data using stl
data_stl_deseason <- data_ts / seasonal(data_stl)     # <<< tutaj sezonowość jest czasami ujemna
plot(data_stl_deseason)



```
```{r forecasts}

data_ts_seasadj <- seasadj(data_stl)
plot(data_ts_seasadj, type="l")
seasonplot(data_ts_seasadj, 12, col=rainbow(25), year.labels=TRUE, main="Seasonal plot: Coal consumption after deseasoning") # seasonal frequency set as 12 for monthly data.


# nie ma dużej różnicy
data_detrend <- lm(data_ts ~ c(1:length(data_ts)))
plot(resid(data_detrend), type="l")  # resid(data_detrend) contains the de-trended series.

# ????
fit <- HoltWinters(data_ts, beta=FALSE, gamma=FALSE)      
fit
f<-forecast(fit, 1)  
plot(forecast(fit, 12), xlab="Year") 

# ETS
fit <- ets(data_ts)
fit
plot(forecast(fit), xlab="Time")
```




```{r czank3}
# outliers ? <<<<<<<<<<<<<<<<<<<<<<<<<<
boxplot.stats(data_ts)$out



```


```{r acf}
acf(data_ts, main="Autocorrelation")         # many spikes outside significant zone - data is NOT RANDOM
pacf(data_ts, main="Partial Autocorrelation")

acf(diff(diff(data_ts, 12), 1), main="Autocorrelation - double diff")          # not many spikes outside significant zone data is RANDOM
pacf(diff(diff(data_ts, 12), 1), main="Partial Autocorrelation - double diff")


library(forecast)
model_arima <- auto.arima(data_ts)

library(tseries)
plot.ts(model_arima$residuals)

# ACF for stationary data
acf(ts(model_arima$residuals), main="ACF Residual")
```

```{r forecast}
forecast_arima <- forecast(model_arima, level=c(95), h=1*12)
plot(forecast_arima)
```

```{r diff}
cbind("Coal consumption" = data_ts,
      "Change in coal" = diff(data_ts, 12),
      "Double diff" = diff(diff(data_ts, 12), 1)
      ) %>% autoplot(facets=TRUE)
```

```{r stationary_test}
library(urca)
data_ts %>% ur.kpss() %>% summary()

diff(data_ts, 12) %>% ur.kpss() %>% summary()

diff(diff(data_ts, 12),1) %>% ur.kpss() %>% summary()

# wystarczy 1 operacja diff, aby otrzymać stationary data
ndiffs(data_ts)

# test pierwiastka jednostkowego

```

```{r test}
Box.test (forecast_arima, lag = 1, type = "Ljung")

```

```{r anomaly}
library(anomalize)
data_tibble <- Lions_Den_data %>% rownames_to_column() %>% as_tibble() %>%
  mutate(date = as.Date(DATE)) %>% select(-one_of('DATE'))

data_tibble %>% time_decompose(`Hard coal consumption per capita [tones]`, method="stl", frequency = "auto", trend = "auto") %>%
  anomalize(remainder, method="iqr", alpha=0.05, max_anoms = 0.2) %>% plot_anomaly_decomposition()

# now for deseasoned data
data_tibble_deseason <- data_deseason %>% as_tibble() %>%
  mutate(date = as.Date(Lions_Den_data$DATE)) %>% select(-one_of('DATE'))

data_tibble_deseason %>% time_decompose(x, method="stl", frequency = "auto", trend = "auto") %>%
  anomalize(remainder, method="iqr", alpha=0.05, max_anoms = 0.2) %>% plot_anomaly_decomposition()



```